---
title: "Data structures I"
output: learnr::tutorial
highlight: null 
mathjax: local
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)

tutorial_options(
   exercise.cap = "Code",
   exercise.completion = FALSE,
   exercise.diagnostics = FALSE,
   exercise.startover = TRUE)
```

<style>
   div.input pre   { color:blue; }
   div.input pre.r { color:red; }
</style>


## Overview

<div style = "background-color:#d6eaf8; padding:10px">
<h4><b>Learning objectives</b></h4> 
In this section, you will learn more about important data structures that we use in R:

* Vectors in more detail: working with character and logical vectors, more sophisticated retrieval of values.
* Factors: a special but important type of vectors that is used to store categorical data.
* Dataframes: creating dataframes, accessing individual columns or rows of a dataframe, importing data into a dataframe, adding new columns to a dataframe.

</div>

<div style = "background-color:#fef9e7; padding:10px"> 
<h4><b>Case</b></h4>  

The company *SmartPack* produces smart backpacks that can be connected to an app on your phone. The app helps to pack your backpack so that it is optimal for your back. The company has an online shop and for every item ordered, the data warehouse records an entry. The data warehouse looks as follows (only the first 5 entries are shown):

```{r, echo=FALSE}
dwh <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  fname = c('Stefanie', 'Martin', 'Elena', 'John', 'John'),
  date = c('2018/07/04', '2018/08/02', '2018/12/22', '2019/03/12', '2019/03/12'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
)
head(dwh)
```

In this section, we will learn how we can explore and work with such data in R.

</div>

<h4><b>Prerequisites</b></h4>
In this section, it is assumed that you know the following R commands: 

* Basic operations such as `+`, `-`, `*`, `/`, etc.
* The assign operator `<-`.
* `c()`, `length()`, `rep()`, `seq()`.
* Accessing values of a vector with `[]` notation.

If your are not familiar with these commands, you are advised to learn the concepts prior to beginning with the current section.

## Vectors in more detail

We have introduced vectors in the **Basic R** tutorial. We learned that there are four different types of vectors: double vectors, integer vectors, character (or string) vectors, and logical vectors. While the **Basic R** tutorial only covered double vectors, here we will see examples of all four types of vectors. We will also learn about some important functions that are helpful when working with vectors.

### Vector types

In the **Basic R** tutorial, we had a look at vectors of type `double`. In the data warehouse of the company *SmartPack* the column with the price would be a vector of type `double`:

<div class = "input">
```{r, comment = ""}
price <- c(350.20, 350.20, 544.10, 210.50, 30.00)
typeof(price)
```
</div>

Another vector type that is also numeric is the `integer` vector. The column `purch_id` (the purchase ID) in the data warehouse looks like it could be a vector of type `integer`. However, we know that R will, by default, assume that any numeric value we enter is a double. Hence, we need to explicitly tell R to transform the vector to integers. We can do that with the function `as.integer()`:

<div class = "input">
```{r, comment = ""}
purch_id <- as.integer(c(21234, 44321, 55437, 95432, 11209))
typeof(purch_id)
```
</div>

Let's check what happens if we transform the vector `price` to integers:

<div class = "input">
```{r, comment = ""}
as.integer(price)
```
</div>

That works! However, we loose the fractional part of the price. For example, the first element is now 350 instead of 350.20. So, generally we do not recommend to transform doubles to integers unless you are sure that there are only whole numbers.

The data warehouse of *SmartPack* also contains character (or string) vectors. For example, the column `lname` contains the last name of the customers.

<div class = "input">
```{r, comment = ""}
lname <- c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen')
typeof(lname)
```
</div>

We can ask R whether or not a vector is a character vector by running the function `is.character()`.

<div class = "input">
```{r, comment = ""}
is.character(lname)
```
</div>

The result of this function is actually a logical vector of length one. It either assumes the value `TRUE` or the value `FALSE`. Let's ask R if `price` is a character vector.

<div class = "input">
```{r, comment = ""}
is.character(price)
```
</div>

The result is correct. R recognizes that `price` is not a character vector and returns `FALSE`. We can also build logical vectors manually:

<div class = "input">
```{r, comment = ""}
y <- c(TRUE, TRUE, FALSE, TRUE)
typeof(y)
```
</div>

Note that logical vectors are especially important when we want to retrieve certain elements of a vector based on a condition (e.g., get all elements that are larger than 10). We will see examples thereof further below.

What happens if we want to concatenate the vector with the prices, `price`, and the vector with the last names, `lname`?

<div class = "input">
```{r, comment = ""}
c(price, lname)
```
</div>

It actually works! However, the prices are now strings and the whole vector is a character vector. This is an example of **coercion**; a concept that causes some people a headache. But it is actually quite simple: R has rules to transform different datatypes to one unique datatype if needed. The most important rule is that R will coerce all doubles, integers or logicals to characters if the vector contains at least one character:

<div class = "input">
```{r, comment = ""}
c(10.75, "hello", TRUE, FALSE, 5L)
```
</div>

### Computing with vectors

In the **Basic R** tutorial, we have seen that `sqrt(x)` will compute the square root of every element in the vector `x`. It is important to know that functions in R perform computations element by element. Let's look at some examples.

<div class = "input">
```{r, comment = ""}
x <- c(2.5, 10, 32, 4.75)
y <- c(10, 12, 22.3, 14.32)
x + y
```
</div>

We see that the elements of `x` and `y` are added elementwise. The same would hold true if we would subtract, multiply or divide the two vectors. The computations would always be performed element by element (try it out for yourself!).

But now you might wonder what happens if the vectors `x` and `y` do not contain the same number of elements. Will R return an error? Or will R have some rules that define how it should cope with such a situation, like in the case of coercion? You probably guessed it, R does have rules for this situation. R will simply reuse elements of the shorter vector until it matches the longer vector. An example might help to understand this:

<div class = "input">
```{r, comment = ""}
x <- c(2.5, 10, 32, 4.75)
y <- c(10, 12, 22.3, 14.32, 33, 2.5, 16, 1)
x + y
```
</div>

Let's have a look at the fifth element of the result. This is the sum of 33 (the fifth element of `y`) and 2.5 (the first element of `x`). Hence, R will in some way double the vector `x` so that the addition can still be performed elementwise. This is important to keep in mind because in some cases this is exactly what you want R to do. In other cases, however, this is not what you want and it might be better if R would return an error. But it won't and you have to be aware of that.

Finally, let's have a look at an example with a character vector. The `substr(x, start, stop)` will return a substring for every element of x depending on the start and stop index. Imagine that for some reason you would like to abbreviate the last name of all your customers by the first two letters of the last name. How can we do that?

<div class = "input">
```{r, comment = ""}
lname <- c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen')
substr(lname, 1, 2)
```
</div>

As you can see, the `substr()` does exactly that. It returns the first two letters of every last name in the vector `lname`. In our data warehouse we could now add this as a new column.

### Accessing elements of a vector and missing values

We have already learned how to use the `[]` notation to access certain elements of a vector. Here, we will learn about some specialties that come with this.

Imagine you try to access an element that is out-of-bound, i.e., you try to access the eighth element of a vector that only contains five elements.

<div class = "input">
```{r, comment = ""}
lname <- c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen')
lname[8]
```
</div>

R will return the value `NA` which stands for *not applicable*. This is R's way of expressing missing or unknown values. We can also set missing values manually. For example, imagine there is a new entry in our data warehouse but we do not know the value of `price` for this entry. We add `NA` manually and then we want to compute the average price that was paid.

<div class = "input">
```{r, comment = ""}
price <- c(price, NA)
mean(price)
```
</div>

This might be a bit surprising to you. Many functions in R will return a missing value if the input vector contains missing values. Fortunately, most functions will allow us to ignore missing values, but you need to explicitly tell the function to do so:

<div class = "input">
```{r, comment = ""}
mean(price, na.rm = TRUE)
```
</div>

In this context, it might be useful to also mention the function `is.na()` which will return a logical vector indicating which elements of a vector are missing.

<div class = "input">
```{r, comment = ""}
is.na(price)
```
</div>

With this we can introduce the concept of logical indexing. Instead of using the index numbers to access elements of a vector (e.g., `price[c(1,3)]`), we can provide a logical vector within the square brackets `[]`. Only the elements where the logical vector is `TRUE` will be returned. For example, if we only want the elements of `price` that are not missing, we can code the following:

<div class = "input">
```{r, comment = ""}
price[!is.na(price)]
```
</div>

Note that `!` is a logical operator that will turn all `FALSE` elements to `TRUE` and vice versa.

Logical indexing is a very powerful concept for retrieving elements of a vector based on one (or many) conditions. For example, imagine we only want to extract prices that are 100 Franks or higher. We can do that with a logical statement within the square brackets:

<div class = "input">
```{r, comment = ""}
price[price >= 100]
```
</div>

Note that all prices that are 100 Franks or larger are returned, but also the `NA` value because R cannot determine whether or not `NA` is larger or equal to 100. If we want to avoid that, we can use the function `which()`. It will return all indices where our condition (`price >= 100`) is `TRUE` and will treat `NA` as `FALSE`:

<div class = "input">
```{r, comment = ""}
price[which(price >= 100)]
```
</div>

The full set of logical operators are:

* `==` for equal to
* `<` and `>` for smaller or larger than
* `<=` and `>=` for smaller or equal to or larger or equal to
* `!=` for not equal to

You will get the chance to practice with those operators in the exercise section.

### Some descriptive statistics with vectors

Imagine you conduct a survey in a company and get a sample of $n=31$ participants with their gender and their hourly wage. The results of this survey can be stored in two vectors:

<div class = "input">
```{r, comment = ""}
gender <- c('m', 'm', 'm', 'm', 'm', 'm', 'f', 'f', 'f', 'f', 'm',
            'm', 'm', 'f', 'm', 'm', 'f', 'm', 'f', 'm', 'm', 'm',
            'm', 'm', 'f', 'm', 'm', 'm', 'f', 'f', 'f')

wage <- c(16, 21, 22, 23, 23, 23, 24, 24, 25, 25, 25,
          25, 25, 26, 26, 26, 27, 27, 28, 28, 28, 29,
          29, 29, 30, 30, 30, 32, 34, 34, 35)
```
</div>

Let's first test whether the vector `wage` really contains 31 elements (you can do the same for `gender`):

<div class = "input">
```{r, comment = ""}
length(wage)
```
</div>

We might be interested how many men and women the sample contains. In R, this is easy:

<div class = "input">
```{r, comment = ""}
table(gender)
```
</div>

We see that the sample contains 11 women and 20 men. Now, in order to get an idea about the wage data, let's print out the most important (univariate) descirptive statistics:

<div class = "input">
```{r, comment = ""}
summary(wage)
```
</div>

Now we got the summary statistics over the full sample. However, we might be more interested in computing descriptive statistics by group. For example, we might wonder whether men are paid more on average than women (based on this sample). For this, we can use the R function `by()`:

<div class = "input">
```{r, comment = ""}
by(wage, gender, mean)
```
</div>

The function `by()` splits the wages into the two groups ("m" and "f") and then computes the mean for each group seperately. In this sample, women have a higher average wage than men. Note however, that we would have to conduct a statistical test in order to determine whether this difference is statistically significant.


### Your turn

```{r, "vec_ex1", exercise = TRUE}
# Combine the double vector a and the integer vector b. What happens?
a <- c(22, 10.5, 3.2)
b <- c(1L, 5L, 3L)
```

```{r "vec_ex1-solution"}
a <- c(22, 10.5, 3.2)
b <- c(1L, 5L, 3L)
c(a, b)
```

```{r, "vec_ex2", exercise = TRUE}
# Divide x by y. What happens?
x <- c(1, 3, 2)
y <- c(1, 2)
```

```{r "vec_ex2-solution"}
x <- c(1, 3, 2)
y <- c(1, 2)
x/y
```

```{r, "vec_ex3", exercise = TRUE}
# s is a character vector. Retrieve a vector with the first letter of each element of s.
s <- c("Hallo", "Hi", "Bonjour", "Ciao")
```

```{r "vec_ex3-solution"}
s <- c("Hallo", "Hi", "Bonjour", "Ciao")
substr(s, 1, 1)
```

```{r, "vec_ex4", exercise = TRUE}
# s is a character vector. Use the function toupper() and see what happens.
s <- c("Hallo", "Hi", "Bonjour", "Ciao")
```

```{r "vec_ex4-solution"}
s <- c("Hallo", "Hi", "Bonjour", "Ciao")
toupper(s)
```

```{r, "vec_ex5", exercise = TRUE}
# The vector lname contains all last names. Modify the code so that only last names starting with 'M' are extracted.
lname <- c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen')
# lname[]
```

```{r "vec_ex5-solution"}
lname <- c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen')
lname[substr(lname, 1, 1) == "M"]
```

### Quiz

```{r, "vec_ex_mc1", echo = FALSE}
question("We have a vector `a <- c(22, 10.5, 3.2)` and a vector `b <- c(FALSE, TRUE, TRUE)`. What is the output of `c(a, b)`?",
  answer("R returns an error. That is not possible."),
  answer("`[1] 22.0 10.5  3.2  0.0  1.0  1.0`", correct = TRUE, message = "This is another example of coercion. The logical value `TRUE` will be coerced to 1.0 and the logical value `FALSE` will be coerced to 0.0. The resulting vector is of type double."),
  answer("`[1] TRUE TRUE TRUE FALSE TRUE TRUE`"),
  answer('`[1] "22.0" "10.5" "3.2" "FALSE" "TRUE" "TRUE"`'),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

```{r, "vec_ex_mc2", echo = FALSE}
question("Given a vector `x <- c(1, 3, 2)`, what is the result of the expression `x^2`?",
  answer("14"),
  answer("`[1] 1 3 2`"),
  answer("4"),
  answer("`[1] 1 9 4`", correct = TRUE, message = "R will compute the square for each element of the vector and return a vector of length 3."),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

## Factors

A factor is a special vector type. Internally, R treats factors as integer vectors and therefore we have not mentioned them above when we were talking about vector types. However, factors are very important in Statistics as they allow us to store categorical (i.e., nominal or ordinal) data, which we often use in statistical data analysis. A factor always has a limited, predefined set of categories (R calls them `levels`). In our data warehouse example, we would transform both the `product` column and the `payment` column to factors. Why? Both of them are limited to a set of different categories that we know in advance. For `product` it is the set of products that *SmartPack* offers. For `payment` it is the set of different payment options that are available in the online shop.

Imagine that *SmartPack* offers five products: four backpack models (Model XX, Model A+, Expert, and Family) and drinking bottles. Let's convert the `product` column into a factor:

<div class = "input">
```{r, comment = ""}
product <- c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle')
product_factor <- factor(product, levels = c('Model XX', 'Model A+', 'Expert', 'Family', 'Drinking bottle'))
```
</div>

Note that we provided all five possible product categories as the `levels` argument, even though the data does not contain the product category "Family".

Let's check the type of the vector:

<div class = "input">
```{r, comment = ""}
typeof(product_factor)
```
</div>

As mentioned before, R treats factors internally as integer vectors. But that is nothing we need to worry about.

Often, categorical data are provided as a so called dummy variable with values 0 and 1. For example, we might have a vector `gender <- c(0, 1, 1, 0, 0, 1, 1, 1, 0)` where 0 denotes men and 1 denotes women. In this case, the levels are `c(0, 1)`, but we might like to add some labels so that we better understand what the values 0 and 1 mean. Let's see how we do that:

<div class = "input">
```{r, comment = ""}
gender <- c(0, 1, 1, 0, 0, 1, 1, 1, 0)
gender_factor <- factor(gender, levels = c(0, 1), labels = c("m", "f"))
gender_factor
```
</div>

R will then print the labels instead of the somewhat abstract levels 0 and 1.

One big advantage of using factors is that many R functions will know what to do when they get a factor as input. For example, the function `table()` will output counts for all categories:

<div class = "input">
```{r, comment = ""}
table(product_factor)
```
</div>

Here, we see why it is a good idea to provide the factor levels (the possible categories) explicitly. Even though, no one bought the "Family" backpack, R still considers this a possible level (with a zero count). Always think carefully about the set of possible levels (categories). If not all of them appear in the data, you will need to provide them explicitly, just the way we did.

You might wonder why we don't convert the columns `lname` and `fname` to factors. Those two columns contain the last name and the first name of the customers, respectively. As mentioned above, to build factors we need to know the set of possible levels in advance. However, in the data warehouse new customers will be added continuously and hence it does not make sense to convert those columns to factors.

### Your turn

```{r, "fac1", exercise = TRUE}
# Above we transformed 'product' into a factor. Now, do the same for 'payment'.
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
```

```{r "fac1-solution"}
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
payment_factor <- factor(payment, levels = c('Credit card', 'Postcard', 'Invoice'))
```

```{r, "fac2", exercise = TRUE}
# Use the code from the first exercise and then count the absolute frequencies per level.
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
```

```{r "fac2-solution"}
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
payment_factor <- factor(payment, levels = c('Credit card', 'Postcard', 'Invoice'))
table(payment_factor)
```

```{r, "fac3", exercise = TRUE}
# Use the code from the first exercise to create the factor.
# Then add a new value 'Credit card' at index 6. What happens?
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
```

```{r "fac3-solution"}
payment <- c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice')
payment_factor <- factor(payment, levels = c('Credit card', 'Postcard', 'Invoice'))
payment_factor[6] <- 'Credit card'
```

### Quiz

```{r, "fac_mc1", echo = FALSE}
question("A factor variable has three predefined levels: 'Credit card', 'Postcard', and 'Invoice'. We add a new value 'Cash' at the end of the factor vector. What happens?",
  answer("R will add an `NA` value because an invalid level has been added.", correct = TRUE, message = "An mentioned above, the set of levels is predefined and if we add new values that are not a valid member of levels, R will insert a missing value."),
  answer("R will correctly add the new value 'Cash' and will create an additional level for it." ),
  answer("R will add the new value but will coerce the factor to a character vector."),
  answer("Nothing happens. R will not add the new value to the factor."),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

## Data frames

Together with vectors, data frames are the most important data structure in R. Whenever you will work with data, it is highly likely that you will need to organize your data as a data frame. Data frames are popular because they allow you to store data of different types (doubles, characters, logicals, etc.) in a rectangular (tabular) format. You can also store different data types in lists, however, lists are less handy to work with and visual data inspection is much easier with data frames. Note that we will not cover lists in this tutorial.

As a matter of fact, the data warehouse we show in the overview of this tutorial is nothing else than a visualization of a R data frame. In this section, we will learn how to create data frames, how to access elements of a data frame (rows, columns, individual values), and how to import data from a file directly into a data frame.

### Creating a data frame

R makes it easy to create your own data frame with the `data.frame()` function and by providing the columns of the data frame as function arguments, e.g.,

<div class = "input">
```{r, comment=""}
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
```
</div>

The `data.frame()` function has the undesired property of treating all character vectors as factors. In the section on factors, we discussed why this may lead to an unwanted outcome. If we add the additional function argument `stringsAsFactors = FALSE`, this behavior can be prevented. With the function `head()`, we can visually inspect the first few rows of our data frame. Let's do that!

<div class = "input">
```{r, comment="", eval = FALSE}
head(df)
```
</div>
<div class = "input">
```{r, comment="", echo = FALSE}
print(head(df))
```
</div>



This works nicely and we get a first idea of how our data looks. Note that the header contains the column names and to the left we see the row names. We did not specify any row names, so by default, rows will be numbered starting from one. Every row corresponds to a data entry referring to the same object (here, every row corresponds to a purchase by a customer). We can get a more technical view of the dataframe using the function `str()` (which stands for structure):

<div class = "input">
```{r, comment=""}
str(df)
```
</div>

A common operation with data frames is to add new columns to already existing data frames. Imagine that the column `price` does not contain the correct price because we granted all our customers a 10% discount. We want to add a column with the actual price that was paid, but we also want to keep the original price. How can we do that? 

<div class = "input">
```{r, comment="", eval = FALSE}
df$price_paid <- df$price * 0.9
head(df)
```
</div>
<div class = "input">
```{r, comment="", echo = FALSE}
df$price_paid <- df$price * 0.9
print(head(df))
```
</div>



We see that adding a new column to the data frame simply requires one line of code. The right side of the assign operator `<-` computes a vector of the new prices based on the old prices. The resulting vector is assigned to the new column `price_paid`. With the `$` sign, we can access one column of the data frame by its column name (more on that in the next subsection).

There are a handful of useful utility functions for data frames that you can test on your own. For example: `is.data.frame()`, `colnames()`, `nrow()`, and `ncol()`.

### Accessing elements of a data frame

To get a specific value in a cell (row-column combination), we first specify the row index and then the column index (separated by a comma). For example, to get the value of `product` (third column) in the second row, we would code:

<div class = "input">
```{r, comment=""}
df[2, 3]
```
</div>

Alternatively, we could access the column via the column name instead of the column index: `df[2, "product"]`.

Sometimes, we need to retrieve the values of one or multiple columns. We can either extract a column as a vector, or we can extract a so called column slice. In the latter case, the result will still be a data frame, whereas in the first case it will be a vector. Let's check it out in practice.

<div class = "input">
```{r, comment=""}
df[["price"]]
```
</div>

Here, we see that the output will be a vector with the prices. Alternative ways to get the same result are `df$price` and `df[ ,"price"]`. Try them out yourself.

We get a column slice if we use single square brackets `[]` instead of the double square brackets `[[]]`. As expected, the result will still be a data frame:

<div class = "input">
```{r, comment=""}
is.data.frame(df["price"])
```
</div>

We can also extract a slice with multiple columns:

<div class = "input">
```{r, comment=""}
df[c("price", "price_paid")]
```
</div>

So far, we have looked at column slices. We can also extract row slices. For example, if we only want to access the first and second row of the data frame, we can use the following code:

<div class = "input">
```{r, comment=""}
df[c(1, 2), ]
```
</div>

Note how we leave the second index specifier after the comma empty. That is, we do not specify which columns to extract and R knows that we want to extract all columns.

One very important tool in our R toolbox is knowing how to extract rows based on conditions of individual values in columns. For example, we want to compute the sum of the paid prices `price_paid` of all customers that paid by credit card.

<div class = "input">
```{r, comment=""}
sum(df[df$payment == "Credit card", "price_paid"])
```
</div>

Let's look at this expression in detail. `df$payment == "Credit card"` will return a logical vector with the value `TRUE` whenever the payment option is "Credit card". We use this logical vector to index the rows (i.e., to only select rows where the payment option was "Credit card"). The second index will select the column we want, namely `price_paid`. The result of this will be a vector of length two which contains the prices paid by the two customers who chose "Credit card" as the payment option. Finally, we simply compute the sum over the elements in this vector with the `sum()` function.

### Importing data into a data frame

So far, we have learned how to create data frames in R and how to access certain elements of a data frame. In practice, we usually want to import data from a CSV or a TXT file directly into a R data frame. R provides many functions that import data from various file formats (even Excel or SPSS files!). Here, we will only learn about the most basic import function in R, `read.table()`.

`read.table()` is a function with many possible arguments (check the help page for more info: `?read.table`). However, we usually only need three function arguments to import our data:

* `file` will specify the file name of the file that should be imported. Note that if we provide only the file name and not a path, R will look for this file in the current working directory (see **Basic R** tutorial for more on working directories).
* `header` will specify whether or not a header line with the column names exists in the file.
* `sep` will specify which separator is used to separate values in the file to be imported.

As mentioned before, we might want to also specify `stringsAsFactors = FALSE` when we import data such that character columns are not automatically transformed into factors.

Imagine you have a CSV (comma-separated values) file named `sales.csv` in your working directory that looks like this:

<div class = "input">
```{r, echo = FALSE, comment = ""}
cat('purch_id;lname;product;price;payment\n21234;Müller;Model XX;350.2;Credit Card\n44321;Sutter;Model XX;350.2;Postcard\n55437;Zuber;Expert;544.1;Credit Card\n95432;Allen;Model A+;210.5;Invoice\n11209;Allen;Drinking bottle;30.0;Invoice')
```
</div>

The file has a header row containing the column names and the values are separated by semicolons. We can import this file into R as follows:

<div class = "input">
```{r, eval = FALSE}
df <- read.table("sales.csv", header = TRUE, sep = ";", stringsAsFactors = FALSE)
```
</div>

In most cases, this simple line of code will be enough to import your file into a R data frame. However, sometimes importing data into R can be surprisingly tedious. A common problem is that R does not handle special characters, such as German Umlauts, properly. In that case, we can check the encoding of our file `sales.csv` and provide it as an argument to the function `read.table()`. For example, if our file is encoded in UTF-8, we would add `fileEncoding = "UTF-8"` to the function.

### Your turn

```{r, "df1", exercise = TRUE}
# Change the column names of the data frame df to whatever names you like.
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
# colnames(df) <- ...
```

```{r "df1-solution"}
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
colnames(df) <- c('id', 'lastName', 'productName', 'amountPaid', 'paymentOption')
```

```{r, "df2", exercise = TRUE}
# Extract the first two rows of the data frame df, but only for the columns 'lname' and 'product'
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
```

```{r "df2-solution"}
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
df[c(1, 2), c("lname", "product")]
```

```{r, "df3", exercise = TRUE}
# Get only entries with a price that is larger than 250 Franks.
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
```

```{r "df3-solution"}
df <- data.frame(
  purch_id = c(21234, 44321, 55437, 95432, 11209),
  lname = c('Müller', 'Sutter', 'Zuber', 'Allen', 'Allen'),
  product = c('Model XX', 'Model XX', 'Expert', 'Model A+', 'Drinking bottle'),
  price = c(350.20, 350.20, 544.10, 210.50, 30.00),
  payment = c('Credit card', 'Postcard', 'Credit card', 'Invoice', 'Invoice'),
  stringsAsFactors = FALSE
)
df[df$price > 250, ]
```

### Quiz

```{r, "df_mc1", echo = FALSE}
question("What are the benefits of using a data frame? (select ALL that apply)",
  answer("It organizes the data in a tabular format.", correct = TRUE),
  answer("It can store data of different data types.", correct = TRUE),
  answer("It is the only data structure that can handle data of different data types."),
  answer("It supports visual inspection of data.", correct = TRUE),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```

```{r, "df_mc2", echo = FALSE}
question("What should we do before we import data from a file? (select ALL that apply)",
  answer("We should empty the workspace of our current R session."),
  answer("We should check if our file has a header line.", correct = TRUE),
  answer("We should check the separator used in our file.", correct = TRUE),
  answer("We should check what the current working directory of our R session is and if our file is really stored there.", correct = TRUE),
  allow_retry = TRUE,
  random_answer_order = TRUE
)
```
